Report for Uniprot-Identified Tissue-Specific Proteins

### 2022-05-23 SeldinReport.Rmd
haven't committed in a while pt2

@@

#setwd("C:/Users/angel/OneDrive/Desktop/Saba Lab/Data/")
setwd("C:/Users/yodeange/OneDrive - The University of Colorado Denver/Saba Lab/Data")
protein_liver = read.csv("Liver_WGCNA.csv")
protein_brain = read.csv("Brain_WGCNA.csv")
secreted_protein_brain = read.csv("Secreted_Brain.csv")
secreted_protein_liver = read.csv("Secreted_Liver.csv")
notsecreted_protein_brain = read.csv("NotSecreted_Brain.csv")
notsecreted_protein_liver = read.csv("NotSecreted_Liver.csv")
sec_annot = read.csv("Sec_Annotation.csv")
tissuespec_protein_brain = read.csv("TissueSpecific_Protein_Brain.csv")
tissuespec_protein_liver = read.csv("TissueSpecific_Protein_Liver.csv")
ts_protein_brain = read.csv("TissueSpecific2_Protein_Brain.csv")
ts_protein_liver = read.csv("TissueSpecific2_Protein_Liver.csv")
tsns_protein_brain = read.csv("NSTissueSpecific_Protein_Brain.csv")
tsns_protein_liver = read.csv("NSTissueSpecific_Protein_Liver.csv")

@@

protein_liver = fix.read.csv(protein_liver)
protein_brain = fix.read.csv(protein_brain)
secreted_protein_brain = fix.read.csv(secreted_protein_brain)
secreted_protein_liver = fix.read.csv(secreted_protein_liver)
notsecreted_protein_brain = fix.read.csv(notsecreted_protein_brain)
notsecreted_protein_liver = fix.read.csv(notsecreted_protein_liver)
sec_annot = fix.read.csv(sec_annot)
tissuespec_protein_brain = fix.read.csv(tissuespec_protein_brain)
tissuespec_protein_liver = fix.read.csv(tissuespec_protein_liver)
ts_protein_brain = fix.read.csv(ts_protein_brain)
ts_protein_liver = fix.read.csv(ts_protein_liver)
tsns_protein_brain = fix.read.csv(tsns_protein_brain)
tsns_protein_liver = fix.read.csv(tsns_protein_liver)

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

THIS WAS DONE WITH TISSUE EXCLUSIVE GENES

@@

#setwd("C:/Users/angel/OneDrive/Desktop/Saba Lab/Data/")
setwd("C:/Users/yodeange/OneDrive - The University of Colorado Denver/Saba Lab/Data")
protein_liver = read.csv("Liver_WGCNA.csv")
protein_brain = read.csv("Brain_WGCNA.csv")
secreted_protein_brain = read.csv("Secreted_Brain.csv")
secreted_protein_liver = read.csv("Secreted_Liver.csv")
notsecreted_protein_brain = read.csv("NotSecreted_Brain.csv")
notsecreted_protein_liver = read.csv("NotSecreted_Liver.csv")
sec_annot = read.csv("Sec_Annotation.csv")
#tissuespec_protein_brain = read.csv("TissueSpecific_Protein_Brain.csv")
#tissuespec_protein_liver = read.csv("TissueSpecific_Protein_Liver.csv")
ts_protein_brain = read.csv("TissueSpecific2_Protein_Brain.csv")
ts_protein_liver = read.csv("TissueSpecific2_Protein_Liver.csv")
tsns_protein_brain = read.csv("NSTissueSpecific_Protein_Brain.csv")
tsns_protein_liver = read.csv("NSTissueSpecific_Protein_Liver.csv")

@@

protein_liver = fix.read.csv(protein_liver)
protein_brain = fix.read.csv(protein_brain)
secreted_protein_brain = fix.read.csv(secreted_protein_brain)
secreted_protein_liver = fix.read.csv(secreted_protein_liver)
notsecreted_protein_brain = fix.read.csv(notsecreted_protein_brain)
notsecreted_protein_liver = fix.read.csv(notsecreted_protein_liver)
sec_annot = fix.read.csv(sec_annot)
#tissuespec_protein_brain = fix.read.csv(tissuespec_protein_brain)
#tissuespec_protein_liver = fix.read.csv(tissuespec_protein_liver)
ts_protein_brain = fix.read.csv(ts_protein_brain)
ts_protein_liver = fix.read.csv(ts_protein_liver)
tsns_protein_brain = fix.read.csv(tsns_protein_brain)
tsns_protein_liver = fix.read.csv(tsns_protein_liver)

@@

#### TSNS ####
# ns tissue-specific -> full
## BRAIN ##

@@

with(compare, qqplot(val[trt == "TSNS brain"], val[trt == "TSS brain"]))
abline(a = 0, b = 1)

@@

with(compare, qqplot(val[trt == "TSNS liver"], val[trt == "TSS liver"]))
abline(a = 0, b = 1)

@@

write.csv(ts_Ssec_brain, "Target_TESec_Ssec_brain.csv")
write.csv(ts_Ssec_liver, "Target_TESec_Ssec_liver.csv")
write.csv(tsns_Ssec_brain, "Target_TENSec_Ssec_brain.csv")
write.csv(tsns_Ssec_liver, "Target_TENSec_Ssec_liver.csv")

@@

with(compare.b, qqplot(val[trt == "TSNS brain"], val[trt == "TSS brain"]))
abline(a = 0, b = 1)

@@

with(compare.l, qqplot(val[trt == "TSNS liver"], val[trt == "TSS liver"]))
abline(a = 0, b = 1)

@@

with(compare.b, qqplot(val[trt == "Not Secreted brain"], val[trt == "Secreted brain"]))
abline(a = 0, b = 1)

@@

with(compare.l, qqplot(val[trt == "Not Secreted liver"], val[trt == "Secreted liver"]))
abline(a = 0, b = 1)

@@

# QQ plot

sec = ts_Ssec_brain$Ssec
nsec = tsns_Ssec_brain$Ssec

cdf = ecdf(sec)

quant = c()
for (i in sec)
{
  gee = cdf(i)
  quant = c(quant, gee)
}

vals.nsec = quantile(nsec, probs = quant)

this = cbind(c(rep("Secreted", times = length(sec)), rep("Not Secreted", times = length(vals.nsec))), c(sec, vals.nsec))
colnames(this) = c("trt", "val")
this = as.data.frame(this)
this$val = as.numeric(this$val)

with(this, qqplot(val[trt == "Not Secreted"], val[trt == "Secreted"]))
abline(a = 0, b = 1)

### 2022-05-23 Sussing Tissue Specific.Rmd
haven't committed in a while pt2

Here we are restricting our protein-coding RNA-seq data to those genes that are labeled as "tissue-specific" on the uniprot database. [Seldin method]. [Definition of tissue-specific on uniprot]. We will calculate Ssec scores for those labeled as tissue-specific on uniprot, comparing secreted to non-secreted. These will be labeled ts uniprot. Additionally, we will calculate these same scores for genes that are labeled tissue-specific AND are only found in the corresponding dataset (brain genes are not found in the liver dataset and vice versa). These will be labeled ts uniprot exclusive. 
Seldin: The Ssec rankings were then filtered ensuring that genes were
primarily expressed by the origin tissue. This was accomplished
by manual inspection of each gene in multi-tissue expression
panels available via BioGPS (Su et al., 2002; Wu et al., 2009) in
both mouse and humans (representative examples using
BioGPS arrays shown in Figure S3). Genes not detected in
BioGPS arrays were investigated for tissue specificity using the
UniProt literature query for ‘‘tissue specificity.’’
This was done after calculation of Ssec matrices. Manual inspection of top genes
For this particular test, it does not appear that they focused on tissue-specific genes at all. Solely focused on the categories secreted and nonsecreted.They tested this with QQPlots
What are BioGPS Arrays?
Would it be useful at all to look at individuals instead of strain means?

## CHECK INF!!!
Do I want to try full dataset as target?

@@

The following analysis is comparing the Ssec distribution between secreted and non-secreted genes. The Wilcoxon-Mann-Whitney test shows a very significant difference in the liver dataset, though with the very small sample size of 8 genes, we cannot be sure of the accuracy of our calculations. The brain dataset continues to not be significantly different between secreted and non-secreted.

@@

@@

wil.brain = coin::wilcox_test(val ~ trt, compare.b)
wil.brain

@@

wil.liver = coin::wilcox_test(val ~ trt, compare.l)
wil.liver

@@

@@

The following analysis is comparing the Ssec distribution between secreted and non-secreted genes. The Wilcoxon-Mann-Whitney test shows a highly significant difference between secreted and non-secreted for the liver dataset. The brain dataset, while not showing a significant p-value, is the closest to significance found in our analyses.

@@

## Looking

```{r}

@@

ggplot(effort, aes(val, color = trt)) + 
  geom_density() +
  labs(title = "Average Reads of Brain and Liver Samples that are Tissue-Specific but NOT Exclusive to Brain/Liver", x = "Ssec", y = "Density")

@@

```
It seems that in the uniprot database, tissue-specicificity does not indicate that a protein is only found in a single tissue, rather that there is SOME KIND of specificity in where it is found.

ENSRNOG00000003732: from top_brain; "Detected in brain (at protein level)"
ENSRNOG00000023337: from top_brain; "Expressed in the dorsal root ganglia"
ENSRNOG00000018962: from top_brain; "Expressed in the ventricle and atrium of adult rats. Also detected in the lung, kidney, liver, skeletal muscle, stomach and urinary bladder. Not detected in brain, colon, testis, spleen or thymus. Overexpressed in the ventricles in the case of hypertension and hypertrophy." in the ensembl database, they say there is expression in brain with an expression score of 56.67, compared to 77.93 in the liver

ENSRNOG00000002811: from the Secreted TS Uniprot Exclusive Brain dataset; "Expressed in olfactory bulb, cerebellum and neuronal layers in hippocampus."

ENSRNOG00000050761: from the Secreted TS Uniprot Exclusive Liver dataset; "Expressed by the liver and secreted in plasma."

ENSRNOG00000012196: from the Secreted TS Uniprot Liver dataset; "Highly expressed in brain, kidney and heart (PubMed:11328816). Expressed at lower level in other tissues such as liver (PubMed:11328816). Expressed in intestine, kidney and liver (at protein level) (PubMed:11328816, PubMed:11330410). Localizes in the epithelia of the jejunum and ileum (PubMed:11330410)."

ENSRNOG00000020889: not exclusive but tissue specific brain; "Minor constituent of the mineralized matrix of bone. Detected in calvaria, rib cartilage, liver, kidney, spleen, brain, lung, skeletal and heart muscle with highest expression in calvaria and approximately half the amount in kidney, liver and brain."

>>>>

Here we are restricting our protein-coding RNA-seq data to those genes that are labeled as "tissue-specific" on the uniprot database.We will calculate Ssec scores for those labeled as tissue-specific on uniprot, comparing secreted to non-secreted. These will be labeled as tissue-specific non-exclusive. Additionally, we will calculate these same scores for genes that are labeled tissue-specific AND are only found in the corresponding dataset (brain genes are not found in the liver dataset and vice versa). These will be labeled tissue-specific exclusive.

@@

@@

ggplot(compare.b, aes(val, color = trt)) + 
  geom_density() +
  labs(title="Brain Ssec Distribution: Tissue-Specific Exclusive Non-Secreted v. Tissue-Specific Exclusive Secreted", x = "Ssec", y = "Density")

@@

wil.brain = coin::wilcox_test(val ~ trt, compare.b)

@@

wil.liver = coin::wilcox_test(val ~ trt, compare.l)

@@

The following analysis is comparing the Ssec distribution between secreted and non-secreted genes that were coded as tissue-specific by Uniprot and determined to be exclusive to brain/liver by direct comparison. The Wilcoxon-Mann-Whitney test shows that neither brain nor liver had significant differences between secreted and non-secreted (brain = `r coin::pvalue(wil.brain)`; liver = `r coin::pvalue(wil.liver)`).

ggplot(compare.l, aes(val, color = trt)) + 
  geom_density() +
  labs(title="Liver Ssec Distribution: Tissue-Specific Exclusive Non-Secreted v. Tissue-Specific Exclusive Secreted", x = "Ssec", y = "Density")

@@

@@

The following analysis is comparing the Ssec distribution between secreted and non-secreted genes that were coded as tissue-specific by Uniprot. This set excluded the extra step of determining exclusivity, so some genes may be shared between the two groups. The Wilcoxon-Mann-Whitney test shows a nearly significant p-value for brain (closest we have gotten so far: `r coin::pvalue(wil.brain)`), but an non-significant p-value for liver (`r coin::pvalue(wil.liver)`).

## Evaluating Non-Exclusive Tissue-Specific Datasets

It seems that in the Uniprot database, tissue-specificity does not indicate that a protein is only found in a single tissue, but rather there is SOME KIND of specificity in where it is found. For example, the following gene is noted as tissue-specific in Uniprot, but was found in the liver dataset:

ENSRNOG00000020889: not exclusive but tissue specific brain; “Minor constituent of the mineralized matrix of bone. Detected in calvaria, rib cartilage, liver, kidney, spleen, brain, lung, skeletal and heart muscle with highest expression in calvaria and approximately half the amount in kidney, liver and brain.”

The following genes were the top genes as selected by the Ssec scores for the brain dataset, but were also found in the liver dataset:

ENSRNOG00000003732: from top_brain; “Detected in brain (at protein level)” 
ENSRNOG00000023337: from top_brain; “Expressed in the dorsal root ganglia” 
ENSRNOG00000018962: from top_brain; “Expressed in the ventricle and atrium of adult rats. Also detected in the lung, kidney, liver, skeletal muscle, stomach and urinary bladder. Not detected in brain, colon, testis, spleen or thymus. Overexpressed in the ventricles in the case of hypertension and hypertrophy.” in the ensembl database, they say there is expression in brain with an expression score of 56.67, compared to 77.93 in the liver

The following graph compares average reads for genes that are tissue-specific and NOT exclusive to brain to those same genes in liver. If tissue-specific annotation from Uniprot is useful, brain should have more reads for these genes than liver, as shown in the graph. The Wilcoxon-Mann_Whitney test affirms this.

@@

ggplot(effort, aes(val, color = trt)) + 
  geom_density() +
  labs(title = "Average Reads of Brain and Liver Samples that are Tissue-Specific but NOT Exclusive to Brain/Liver", x = "Avg Num Reads", y = "Density")

@@
